Unbiased CVM Algorithm

### 2024-12-16 cvm_unbiased/main.tex
fix some minor typos

\begin{algorithm}
\caption{Unbiased CVM algorithm}
\begin{algorithmic}
\State $\chi \gets \emptyset$; $p \gets 1$
\For{$i \gets 1$ to $l$}
  \State $\chi \gets \chi - \{a_i\}$
  \State $\mathrm{coin} \gets \mathrm{Ber}(p)$
  \If{$\mathrm{coin}$}
    \State $\chi \gets \chi \cup \{a_i\}$
  \EndIf
  \If{$|\chi| = n$}
    \State $\chi \gets U(\{ S \subseteq \chi \,|\, |S|=nf \})$
    \State $p \gets p f$
  \EndIf
\EndFor
\State \Return{$p^{-1} |\chi|$}
\end{algorithmic}
\end{algorithm}
We will denote the first five lines of the loop as step 1, the last four lines as step 2, and the distribution of the state of the algorithm after processing $i$ elements of the sequence by $\Omega_i$.
The elements of the probability spaces are pairs composed of a set and a fraction, representing $\chi$ and $p$. 
For example: $\Omega_0 = U(\{(\emptyset, 1)\})$ is the initial state, $\Omega_1 = U(\{(\{a_1\}, 1)\})$, etc. $\Omega_l$ denotes the final state. 
We introduce $\chi$ and $p$ as random variables defined over such probability spaces $\Omega$, in particular, $\chi$ (resp. $p$) is the projection to the first (resp. second) component.

@@

We distinguish the two cases $a_{k+1} \in S$ and $a_{k+1} \notin S$.
\\
If $a_k \in S$ then:
\begin{eqnarray*}
  \expectation_{\Omega’_{k+1}}\left[\frac{I(S \subseteq \chi)}{p^{|S|}}\right] & = & \int_{\Omega_k} \frac{I(S' \subseteq \chi(\omega))}{ p(\omega)^{|S'|+1}}
  \int_{\mathrm{Ber}(p(\omega))} I(\tau) \, d \tau \, d \omega \\ 
  & = & \int_{\Omega_k} p(\omega)^{-(|S'|+1)} I(S' \subseteq \chi(\omega)) p(\omega) \, d \omega \\
  & = & \expectation_{\Omega_k}\left[p^{-|S'|} I(S' \subseteq \chi)\right] \leq_{IH} 1
\end{eqnarray*}
If $a_k \notin S$ then:
\begin{eqnarray*}
  \expectation_{\Omega’_{k+1}}\left[\frac{I(S \subseteq \chi)}{p^{|S|}}\right] & = & \int_{\Omega_k} \frac{I(S' \subseteq \chi(\omega))}{p(\omega)^{|S'|+1}}
  \int_{\mathrm{Ber}(p(\omega))} 1 \, d \tau \, d \omega \\ 
  & = & \int_{\Omega_k} p(\omega)^{-(|S'|)} I(S' \subseteq \chi(\omega)) \, d \omega \\
  & = & \expectation_{\Omega_k}\left[p^{-|S'|} I(S' \subseteq \chi)\right] \leq_{IH} 1
\end{eqnarray*}
\paragraph{Induction step $Q(k+1) \rightarrow P(k+1)$:} \phantom{.}\\
Let $S \subseteq \{ a_1, \ldots, a_{k+1} \}$.
Let us also introduce the abbreviation $C(X) := \{T \subset X \, | \, |T| = n f \}$, i.e., the set of $nf$-subsets of $X$.

@@

To get a tail estimate, we use the Cram\'{e}r-Chernoff method: 
{\allowdisplaybreaks
\begin{eqnarray*}
  \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A| \right) & \leq & \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A'| \right) \\
  & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
 & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} \exp ( t \tilde{p}^{-1} I(s \in \chi) ) \right] \\  
  & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} (1 + I(s \in \chi) (e^{t \tilde{p}^{-1}}-1) ) \right] \\  
 & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} \left(1 + \frac{a(t)}{p} I(s \in \chi) \right) \right] \\  
 & = & \exp( -t (1+\eps) |A| ) \sum_{T \subseteq A'} a(t)^{|T|} \expectation_\Omega \left[ \prod_{s \in T} (p^{-1} I(s \in \chi) ) \right] \\  
 & \underset{Lemma~\ref{le:neg_cor}}{\leq} & \exp( -t (1+\eps) |A'| ) \sum_{T \subseteq A'} a(t)^{|T|} \\  
 & = & \left(\exp( -t (1+\eps) |A| ) (1+a(t))\right)^{|A'|} \\ & \leq & \exp ( |A| (\ln(1+a(t))-t (1+\eps)))
\end{eqnarray*}}
So we just need to show that (using $|A|=\frac{n}{4q}$) for some $t > 0$:
\[
  \ln(1+a(t))-t (1+\eps) \leq \frac{-q \eps^2}{3}
\]
which can be verified for $t = q \ln ( 1 + \frac{\eps}{1-q(1+\eps)}) > 0$.
The latter can be established by analyzing the function $f(q,\eps) = -\ln (1+a(t)) +t (1+\eps) - \frac{q\eps^2}{3}$.
For which it is easy to check $f(q,0) = 0$ and the derivative with respect to $\eps$ is positive in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e.,
$f(q,\eps) \geq 0$.
\end{proof}

@@

First, we bound the probability of jumping from $p=f^j$ to $p=f^j+1$ at a specific point in the algorithm, e.g., after processing $k$ stream elements.
It can only happen if $|\chi|=n$, $p=f^j$ in $\Omega'_k$.

>>>>

\begin{algorithm}
\caption{Unbiased CVM algorithm}
\begin{algorithmic}[1]
\State $\chi \gets \emptyset$; $p \gets 1$
\For{$i \gets 1$ to $l$}
  \State $\chi \gets \chi - \{a_i\}$
  \State $\mathrm{coin} \gets \mathrm{Ber}(p)$
  \If{$\mathrm{coin}$}
    \State $\chi \gets \chi \cup \{a_i\}$
  \EndIf
  \If{$|\chi| = n$}
    \State $\chi \gets U(\{ S \subseteq \chi \,|\, |S|=nf \})$
    \State $p \gets p f$
  \EndIf
\EndFor
\State \Return{$p^{-1} |\chi|$}
\end{algorithmic}
\end{algorithm}

We will denote the first five lines of the loop ($4$--$7$) as step 1, the last four lines ($8$--$11$) as step 2, and the distribution of the state of the algorithm after processing $i$ elements of the sequence by $\Omega_i$.
The elements of the probability spaces are pairs composed of a set and a fraction, representing $\chi$ and $p$.
For example: $\Omega_0 = U(\{(\emptyset, 1)\})$ is the initial state, $\Omega_1 = U(\{(\{a_1\}, 1)\})$, etc. $\Omega_l$ denotes the final state.
We introduce $\chi$ and $p$ as random variables defined over such probability spaces $\Omega$, in particular, $\chi$ (resp. $p$) is the projection to the first (resp. second) component.

@@

We distinguish the two cases $a_{k+1} \in S$ and $a_{k+1} \notin S$.\\

If $a_{k+1} \in S$ then:
\begin{eqnarray*}
  \expectation_{\Omega’_{k+1}}\left[\frac{I(S \subseteq \chi)}{p^{|S|}}\right] & = & \int_{\Omega_k} \frac{I(S' \subseteq \chi(\omega))}{ p(\omega)^{|S'|+1}}
  \int_{\mathrm{Ber}(p(\omega))} I(\tau) \, d \tau \, d \omega \\
  & = & \int_{\Omega_k} p(\omega)^{-(|S'|+1)} I(S' \subseteq \chi(\omega)) p(\omega) \, d \omega \\
  & = & \expectation_{\Omega_k}\left[p^{-|S'|} I(S' \subseteq \chi)\right] \leq_{IH} 1
\end{eqnarray*}

If $a_{k+1} \notin S$ then:
\begin{eqnarray*}
  \expectation_{\Omega’_{k+1}}\left[\frac{I(S \subseteq \chi)}{p^{|S|}}\right] & = & \int_{\Omega_k} \frac{I(S' \subseteq \chi(\omega))}{p(\omega)^{|S'|+1}}
  \int_{\mathrm{Ber}(p(\omega))} 1 \, d \tau \, d \omega \\
  & = & \int_{\Omega_k} p(\omega)^{-(|S'|)} I(S' \subseteq \chi(\omega)) \, d \omega \\
  & = & \expectation_{\Omega_k}\left[p^{-|S'|} I(S' \subseteq \chi)\right] \leq_{IH} 1
\end{eqnarray*}
\paragraph{Induction step $Q(k+1) \rightarrow P(k+1)$:} \phantom{.}\\
Let $S \subseteq \{ a_1, \ldots, a_{k+1} \}$.
Let us also introduce the abbreviation $C(X) := \{\tau \subset X \, | \, |\tau| = n f \}$, i.e., the set of $nf$-subsets of $X$.

@@

To get a tail estimate, we use the Cram\'{e}r-Chernoff method:
{\allowdisplaybreaks
\begin{eqnarray*}
  \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A| \right) & \leq & \prob_{\Omega} \left( \tilde{p}^{-1} |\chi| \geq (1+\eps) |A'| \right) \\
  & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \exp ( t \tilde{p}^{-1} |\chi| ) \right] \\
 & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} \exp ( t \tilde{p}^{-1} I(s \in \chi) ) \right] \\
  & = & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} (1 + I(s \in \chi) (e^{t \tilde{p}^{-1}}-1) ) \right] \\
 & \leq & \exp( -t (1+\eps) |A| ) \expectation_\Omega \left[ \prod_{s \in A'} \left(1 + \frac{a(t)}{p} I(s \in \chi) \right) \right] \\
 & = & \exp( -t (1+\eps) |A| ) \sum_{T \subseteq A'} a(t)^{|T|} \expectation_\Omega \left[ \prod_{s \in T} (p^{-1} I(s \in \chi) ) \right] \\
 & \underset{Lemma~\ref{le:neg_cor}}{\leq} & \exp( -t (1+\eps) |A| ) \sum_{T \subseteq A'} a(t)^{|T|} \\
 & = & \left(\exp( -t (1+\eps) |A| ) (1+a(t))\right)^{|A'|} \\ & \leq & \exp ( |A| (\ln(1+a(t))-t (1+\eps)))
\end{eqnarray*}}
So we just need to show that (using $|A|=\frac{n}{4q}$) for some $t > 0$:
\[
  \ln(1+a(t))-t (1+\eps) \leq \frac{-q \eps^2}{3}
\]
which can be verified for $t = q \ln ( 1 + \frac{\eps}{1-q(1+\eps)}) > 0$.
The latter can be established by analyzing the function $f(q,\eps) = -\ln (1+a(t)) +t (1+\eps) - \frac{q\eps^2}{3}$.
For which it is easy to check $f(q,0) = 0$ and the derivative with respect to $\eps$ is positive in the range $0 \leq q \leq 1/4$ and $0 < \eps \leq 1$, i.e.,
$f(q,\eps) \geq 0$.
\end{proof}

@@

First, we bound the probability of jumping from $p=f^j$ to $p=f^{j+1}$ at a specific point in the algorithm, e.g., after processing $k$ stream elements.
It can only happen if $|\chi|=n$, $p=f^j$ in $\Omega'_k$.
